---
title: "Recovery rate for standards with kaolin"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(rjags)
library(stringr)

library(here)


# Default ggplot theme
theme_set(
  theme_bw() +
  theme(text = element_text(size = 20))
)

loadFrom <- function(file) {
  e <- new.env()
  load(file, envir = e)
  
  o <- ls(e)
  if (length(o) == 1) get(o, envir = e)
  else as.list(e)
}


set.seed(42)

```

## Summary

There are no data for the carbon content of undigested standards of charcoal plus kaolin. In the following analysis we assume that these missing data follow the same distribution as carbon content for undigested standards without kaolin. We then impute values from this distribution and compare them to values for digested standards with kaolin to estimate recovery rate.

## Data

```{r}

predigest <- loadFrom( here("data/predigest.RData") )

digest <- loadFrom( here("data/digest.RData") )

```



## JAGS model code to fit distributions and estimate recovery rate

```{r}

modelTxt <- "
model {
  # undigested samples
  for (i in 1:NPre) {
    xpre[i] ~ dt(muPre, tauPre, shapePre)
  }
  
  # digested samples
  for (i in 1:NDig) {    
    xdig[i] ~ dt(muDig, tauDig, shapeDig)
  }

  muDig <- recoveryRate * muPre
  
  muPre ~ dnorm(0, 0.0001)
  sdPre ~ dunif(0.0, 10)
  tauPre <- 1 / (sdPre * sdPre)
  shapePre <- shapePreMinusOne + 1
  shapePreMinusOne ~ dexp(1/29)

  sdDig ~ dunif(0.0, 10)
  tauDig <- 1 / (sdDig * sdDig)
  shapeDig <- shapeDigMinusOne + 1
  shapeDigMinusOne ~ dexp(1/29)
  
  recoveryRate ~ dbeta(1, 1)
}"

```


## Function to fit model

This function fits the nodel for a selected standard identified by label. It then checks the model for convergence and, if all is well, returns posterior samples as a data frame.

```{r}

fit_model <- function(label, n.iter = 10000) {
  dat.undigested <- filter(predigest, newLabel == label)
  dat.digested <- filter(digest, newLabel == paste0(label, "_kaolin"))
  
  zz <- textConnection(modelTxt)
  
  model <- jags.model(
    file = zz,
    
    data = list(
      NPre = nrow(dat.undigested),
      xpre = dat.undigested$percentC,
      NDig = nrow(dat.digested),
      xdig = dat.digested$percentC),
    
    inits = function() {
      list(muPre = 0, sdPre = 1, shapePreMinusOne = 29,
           sdDig = 1, shapeDigMinusOne = 29,
           recoveryRate = runif(1))
    },
    
    n.chains = 2,
    
    quiet = TRUE
  )
  
  close(zz)
  
  update(model, n.iter)
  
  sims <- coda.samples(model, 
                       c("recoveryRate", 
                         "muPre", "sdPre", "shapePre", 
                         "sdDig", "shapeDig"), n.iter=n.iter)
  
  # check model convergence
  g <- gelman.diag(sims)
  m <- g$psrf
  
  if ( !all(m[, 1] - 1.0 < 0.1) )
    stop("model did not converge for label: ", label, "\n",
         m)

  # If all is well return the results of all chains combined
  as.data.frame( do.call(rbind, sims) )
}

```


## Estimate recovery rate for selected standards

Fit the model for each standard label, check that the model converged, then report on recovery rate.

```{r}

labels <- c("A", "B", "C", "D", "charcoal_1%", "charcoal_5%")

fits <- lapply(labels, fit_model, n.iter = 20000)

rates <- lapply(fits, 
                function(sims) {
                  with(sims,
                       data.frame(
                         lwr = quantile(recoveryRate, 0.025),
                         mean = mean(recoveryRate),
                         median = quantile(recoveryRate, 0.5),
                         upr = quantile(recoveryRate, 0.975)
                       )
                  )
                })

rates <- do.call(rbind, rates)
rownames(rates) <- NULL
rates$label <- labels

rates <- select(rates, label, everything())

knitr::kable(rates, caption = "Estimated recovery rates", digits = 2)

```


